---
title: "EDA"
author: "Particle Clustering Proj Group"
date: "`10/17/2025`"
output: pdf_document
---

```{r}
library(tidyverse)

train_data <- read_csv("data-train-processed.csv")


#Fix the infinity values in Fr via inverse logit transformation, but keep Fr as numeric
train_data$Fr <- as.numeric(tolower(as.character(train_data$Fr)))
train_data$Fr_invlogit <- 1 / (1 + exp(-train_data$Fr))


```

```{r}
library(GGally)
ggpairs(
  train_data |> select(St, Re, Fr_invlogit, mean, variance, skewness, kurtosis)
) 
```

```{r}
#check individual distributions of variables for more detail 
hist(train_data$St, breaks=20)
hist(train_data$Re, breaks=20)
hist(train_data$Fr_invlogit, breaks=20)
hist(train_data$mean, breaks=20)
hist(train_data$variance, breaks=20)
hist(train_data$skewness, breaks=20)
hist(train_data$kurtosis, breaks=20)
```

Immediate Notes

-   Collinearity between the 3 predictors is not a large concern

-   log transformation on mean/var/skewness/kurtosis will be very helpful for reducing the impact of right skew on the response variable. Recommended log transform on St and (potentially Fr_invlogit).

```{r}

basicmean_lm <- lm(mean ~ St + Re + Fr_invlogit, data = train_data)
basicvar_lm  <- lm(variance ~ St + Re + Fr_invlogit, data = train_data)
basicskew_lm <- lm(skewness ~ St + Re + Fr_invlogit, data = train_data)
basickurt_lm <- lm(kurtosis ~ St + Re + Fr_invlogit, data = train_data)

summary(basicmean_lm)
summary(basicvar_lm)
summary(basicskew_lm)
summary(basickurt_lm)

```

Observations: mean performs well as just a base model. The rest are subpar to okay going off of R\^2.

### 
