---
title: "Predicting Particle Cluster Distributions for Turbulence Flow"
author: "FILL"
output: pdf_document
---

#DELETE LATER

-   understand data, what to do with "inf"

-   calculate 4 new cols, one for each metric we predict (mean, sd, skew, kurt)

-   conduct EDA (ggpairs, look for nonlinear relationships and correlated predictors)

-   for each of the 4 categories (but apply same method across alll 4): MUST DECIDE METHOD: which model do we start with (full interactions?)? How do we evaluate the best model selection methods (BIC/AIC, ANOVA, CV, subset/fwd/bwd selection....)? How do we select what type of model to use (OLS, ridge, lasso)? ———hardest part (bc code should be straightforward)

-   run models through testing data and save into new csv

-   finalize all 4 models, and interpret coefficients (results) -\>provide insights into broader research question in conclusion

# Introduction

# Methods

```{r}
training_data <- read.csv("data-train-processed.csv")
training_data
```

```{r linear reg}

linear_model <- lm(mean ~ St + Re + Fr, data = training_data)

summary(linear_model)

#yeah idk about the inf values
```

# Results

## 3. Interpretation of the Skewness Model

The final skewness model that was selected was:

$$
\log(\text{skewness}) \sim Fr_{\text{invlogit}} + Re + Re:Fr_{\text{invlogit}}
$$

This model explains approximately **55%** of the variation in log-skewness ($R^2 = 0.55$), suggesting a moderate fit that captures key physical trends while leaving room for turbulence-driven variability.\
Residual and Q–Q plots indicate approximate normality with some heteroscedasticity—reasonable given that skewness (a third-order statistic) tends to fluctuate strongly with turbulence intensity.

------------------------------------------------------------------------

### **Coefficient Interpretation**

Three predictors were retained: **Froude number (**$Fr_{\text{invlogit}}$), **Reynolds number (**$Re$), and their **interaction (**$Re \times Fr_{\text{invlogit}}$).

-   **Froude number (**$Fr_{\text{invlogit}}$):\
    As $Fr$ increases (weaker gravity effects), skewness decreases, meaning particle cluster distributions become more symmetric.\
    When gravity dominates (low $Fr$), particles settle downward, forming more elongated and right-skewed clusters.

-   **Reynolds number (**$Re$):\
    Higher turbulence (larger $Re$) reduces skewness, as intense mixing disrupts organized, gravity-driven cluster structures.\
    Low $Re$ flows allow coherent formations that increase asymmetry.

-   **Interaction (**$Re \times Fr_{\text{invlogit}}$):\
    The impact of gravity depends on turbulence intensity.\
    At low $Re$, gravity strongly drives asymmetric clustering, but at high $Re$, turbulence disperses particles enough to weaken this effect.

Because the model operates on $\log(\text{skewness})$, all effects are **multiplicative** on the original scale.\
For instance, a coefficient of $+0.3$ corresponds to a $\approx 35\%$ increase in actual skewness ($e^{0.3} \approx 1.35$).

### **Model Selection**

Cross-validation favored **LASSO** over **OLS** by a narrow margin (CV RMSE: 0.779 vs. 0.784), reflecting a slight reduction in variance without significant bias increase.\
The near-identical errors indicate that both models capture similar structure, but **LASSO’s regularization** improves stability for this noisy, higher-moment response.

### **Prediction Intervals and Scientific Meaning**

The **95% prediction intervals** are wide, primarily due to limited Reynolds number levels (90, 224, 398) and stochastic variation in turbulence.\
Still, the model provides physically meaningful directional trends:

-   **High** $Re$: reduces skewness through enhanced turbulent mixing.
-   **Low** $Fr$: increases skewness via stronger gravitational clustering.
-   $St$ (Stokes number): excluded from selection, suggesting particle inertia adds minimal explanatory power once $Re$ and $Fr$ are accounted for.

Overall, the model shows that **turbulence homogenizes particle distributions**, while **gravity induces asymmetry**. Although uncertainty is substantial, the observed relationships align with theoretical expectations for particle clustering in turbulent flow environments.

# Conclusions
